---
title: "Reproducible Research: Peer Assessment 1"
author: rudini1
date: "15.12.2017"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

## Loading and preprocessing the data

The file can be downloaded at the specified site. Alternatively the file is also located on the GitHub repository cloned.
```{r message=FALSE, warning=FALSE}
if(!file.exists("activity.csv")){
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", "activitymonitoringdatafile.zip")
unzip("activitymonitoringdatafile.zip")
}
library(data.table)
library(lubridate)
library(dplyr)
library(ggplot2)

# Sys.setlocale("LC_ALL","En") # this is necessary for the weekdays to be in english on a non-english system
# Sys.setenv(LANG = "en")

activity <- fread("activity.csv", header = TRUE, sep = ",", na.strings = "NA") %>%
mutate(date = as.POSIXct(date, "%Y-%m-%d"))

```


## What is mean total number of steps taken per day?

```{r}

totalstepsperday <- activity %>%
      group_by(date) %>%
      summarise(totalsteps = sum(steps, na.rm = TRUE))

# ggplot(data = totalstepsperday, aes(x =  date, y= totalsteps)) + geom_col()
ggplot(data = totalstepsperday, aes(x = totalsteps)) + 
      geom_histogram(na.rm = TRUE) +
      theme_bw()

meanstepsperday <- round(mean(totalstepsperday$totalsteps), 0)
medianstepsperday <- median(totalstepsperday$totalsteps)
```

 - The median of the steps per day is : `r medianstepsperday`
 - The mean of the steps per day is : `r meanstepsperday`

## What is the average daily activity pattern?

Average activity vs. time in the day. The time is given in the interval column as minutes after midnight

```{r}
averagestepsperinterval <- activity %>% 
      group_by(interval) %>% 
      summarise(meanstepsperinterval = mean(steps, na.rm = TRUE))

ggplot(data = averagestepsperinterval, aes(x = seq(from = 0, by = 5/60, along.with = interval), y = meanstepsperinterval)) + 
      geom_line() +
      # geom_smooth(method = "loess", span = 0.05, se = FALSE) +
      theme_bw() +
      labs(x = "average day (h)", y = "average steps over all days")

```

```{r echo=TRUE, results="hide"}
      maxsteps <- averagestepsperinterval[which.max(averagestepsperinterval$meanstepsperinterval),]$interval
```

The interval with maximum average steps is : `r maxsteps` what is 8:35 in the morning

## Imputing missing values

first find the number of rows with missing (NA) values. This is equal to the number of rows that are not complete

```{r}
missingValues <- sum(!complete.cases(activity))
```
there are `r missingValues` rows with missing values (from a total of `r dim(activity)[1]`)

imputing missing values: strategy is to use the average steps of the intervals computed in the previous steps
so for missing days, the average corresponding to the time interval is imputed

```{r}
activity %>% 
  select_if(function(x) any(is.na(x))) %>%  # select the columns with NA
  summarise_each(funs(sum(is.na(.))))  # sum the NA's on the selected columns
```
only the variable "steps" contains missing values
the next code block imputes the missing values in variable "steps"
```{r}
nas <- which(is.na(activity$steps))
corrs <- match(activity$interval, averagestepsperinterval$interval) # same as rep(1:288, times = 61)
activityImputed <- activity # first copy all data 
activityImputed[nas,1] <- averagestepsperinterval[corrs[nas],2] # automatic coercion of steps form int to num 
```

next code makes a histogram of the total steps per day as in the beginningm, but with the imputed dataset
```{r}
totalstepsperdayimputed <- activityImputed %>%
      group_by(date) %>%
      summarise(totalsteps = sum(steps, na.rm = TRUE))

# ggplot(data = totalstepsperdayimputed, aes(x =  date, y= totalsteps)) + geom_col()
ggplot(data = totalstepsperdayimputed, aes(x = totalsteps)) + 
      geom_histogram(na.rm = TRUE) +
      theme_bw()

meanstepsperdayimp <- round(mean(totalstepsperdayimputed$totalsteps), 0)
medianstepsperdayimp <- median(totalstepsperdayimputed$totalsteps)
```

There were days full of NA's giving a large number of days with zero steps. so the histogram differes in the nummer of days with zero steps when using imputed set.

 - The median of the steps per day is : `r sprintf("%.0f", medianstepsperdayimp)`
 - The mean of the steps per day is : `r sprintf("%.0f", meanstepsperdayimp)`  
 
The distribution is now much more symmetric with both, median and mean beeing the same and different from the non-imputed data


## Are there differences in activity patterns between weekdays and weekends?

```{r}
# Definitions
weekend <- c("Samstag","Sonntag")  # I had to use the german words for saturday and Sunday here you have to use the words in your locale. I didn't find how to change it to english locale on my system

# weekday v. weekend
activityImputed <- activityImputed %>% 
      mutate(isweekend = if_else(weekdays(date) %in% weekend, "weekend", "weekday")) %>% 
      mutate(isweekend = as.factor(isweekend))
# table(activityImputed$isweekend)


stepsWeek <- activityImputed[activityImputed$isweekend == "weekday",]
stepsWeekend <- activityImputed[activityImputed$isweekend == "weekend",]

totalstepsperworkday <- stepsWeek %>% group_by(interval) %>%
      summarise(avgsteps=mean(steps, na.rm = TRUE))

totalstepsperweekendday <- stepsWeekend %>% group_by(interval) %>%
      summarise(avgsteps=mean(steps, na.rm = TRUE))
```

Plot the data
```{r}
par(mfrow=c(2,1),
    mar = c(1,1,1,1),
    oma = c(1,1,1,1))
with(totalstepsperweekendday, 
     plot(interval, avgsteps, type = "l", xlim = c(0,2400), ylim = c(0,250)))
title("average number of steps on weekends", line = -1)
with(totalstepsperworkday, 
     plot(interval, avgsteps, type = "l", xlim = c(0,2400), ylim = c(0,250)))
title("average number of steps on workdays", line = -1)

```

we see, that on workdays, the activity peaks higher around 8:30 than on weekends. However, on weekends activity is generally higher around noon and in the afternoon


